\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-07-31 docs/parco13/paper.tex
More text changes

\begin{abstract}
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ with $m^d$ degrees of freedom, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to integrate with the \pforest AMR library since it allows us to compose
the mesh from multiple mapped octrees and enables the cubed sphere and other
nontrivial multiblock geometries.  We describe aspects of the parallel
implementation and close with scalings for both MPI-only and OpenMP/MPI hybrid
runs, where the largest MPI-only run uses 4,096 CPU cores.
\end{abstract}

@@

With MPI parallelism, neighbor patches can be either local or assigned to a
different process (ghost patches).  Since this must make no difference
numerically, we need to ensure that the values of ghost patches are current
whenever they are accessed.  To this end, we allocate storage for a layer of
ghost patches in \forestclaw, which we can pass to a general-purpose \pforest
routine that communicates local leaf data to all processes that view a
particular leaf as a ghost.  If we call this routine before we go through the
local neighbor interactions, we can handle ghost values implicitly by neighbor
lookups without querying if they are local or remote.

@@

The threaded parallelism over the degrees of freedom of a patch can be handled
by \forestclaw alone without the need to involve \pforest.  Additionally,
many-core implementations, such as \manyclaw \cite{manyclaw}, can be used for
the integration of the hyperbolic system on a leaf-patch thereby allowing for
hybrid parallelism.  The design of leaf-patches then can enable efficient
management of data local to many-core architectures (for example the Xeon Phi
coprocessor) and the host.

@@

In both sets of numerical results, we solve a scalar advection
equation,
\begin{equation}
q_t + ({\bf u}\cdot \nabla) q = 0 ,
\end{equation}
where the velocity field ${\bf u} = (u(\xi,\eta,t), v(\xi,\eta,t)$ is
a prescribed function of computational coordinates.  The relevant
numerical parameters that are set in each case include the size of the
patch on each leaf ($m=8$ or $m=16$), and the minimum and maximum
refinement levels, which in turn fix minimum and maximum effective mesh
resolutions.

@@

\begin{description}
\item{\bf Example 1:} An initial concentration field $q$
is set to 0 in the left half of a computational square and to 1 in the
right half.  A time dependent flow field is prescribed that distorts
the interface between the 0 and 1 concentration values.  The minimum
and maximum levels of refinement are set to 3 and 6, respectively.
This results in a minimum resolution of $64 \times 64$ and a maximum
resolution of $512 \times 512$.  \Fig{results2d} shows the results
at an intermediate time step.

@@

\item{\bf Example 2:} We demonstrate the
multiblock functionality of \forestclaw by considering flow on a
sphere.  The sphere mapping we use consists of two quadtrees, each
defined in the computational space $[-1,1] \times [-1,1]$
\cite{ca-he-le:2008, be-ca-he-le:2009}.  Each quadtree is mapped to cover one
hemisphere.  The initial conditions is $q = 0$ in one half of
the sphere, and $q = 1$ in the other half, where the halves are not
necessarily aligned with the equator of the mapping.  The flow field
${u}$ simulates rigid body rotation.  We show the results at an intermediate
time in \Fig{results2d}.
\end{description}

\begin{figure}
\begin{center}
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_advance_strong-crop}
  \\[1ex]
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_exchange_strong-crop}
\end{center}
\caption{Strong scaling of MPI parallelism for advection on a
  uniformly refined mesh.  Top: run time of time integration.  Bottom:
  time spent in local and parallel neighbor exchanges.  The number of
  patches is $2^{2 \ell}$ at level $\ell$.  The flat lines for the
  smaller runs are caused by having fewer patches than MPI processes,
  leaving some of the processors idle.  For each data point we use the
  maximum time required across all processes. \comment{This last
    sentence is a bit unclear.  What do you mean exactly?}}
\label{fig:uniscale}
\end{figure}%
We begin our analysis of Example 1 with uniformly-refined mesh experiments run on TACC's
Stampede supercomputer at different levels of resolution.
We use 1 to 16 MPI processes on one 16-core compute node, obtaining a strong
scaling efficiency for the time integration between 97\%--103\% depending on
the resolution (from levels 1 and 8).  Then, we examine processor counts from
16 to 256 using 16 cores per node.  The scaling behavior remains linear for
the time integration itself, and mostly linear for the local and parallel
neighbor exchanges which require about 10\% of the time integration run time;
see \Fig{uniscale}.

We have examined weak scalability as well by using four times as many processes
for each level increase, from level 6 at 64 cores to level 10 at 4096.  The
exchange times stay consistently below 16\% of the time integration.  The time
in seconds per time integration step varies between .0578 and .0602,
which yields a parallel efficiency of 96\%.

\begin{table}
  \begin{center}
    \begin{tabular}{cccc|rr}
      \multicolumn{4}{c|}{strategy} & \multicolumn{2}{c}{wall clock time} \\
      mesh & remesh & partition & time step & $P = 16$ & $P = 256$ \\
      \hline
      uniform & none       & by count  & global   & 3961.  & 256.  \\
      \hline
          AMR & every step & by count  & global   &  252.  &  54.6 \\
          AMR & every 4    & by count  & global   &  178.  &  39.7 \\
      \hline
          AMR & every step & by count  & subcycle &   99.9 &  17.3 \\
          AMR & every 4    & by count  & subcycle &   87.2 &  14.0 \\
      \hline
          AMR & every step & by weight & subcycle &   95.7 &  18.2 \\
          AMR & every 4    & by weight & subcycle &   84.4 & 14.2 \\
      \hline
    \end{tabular}
  \end{center}
\caption{Comparison of different meshing strategies for the advection example
  with a fixed maximum level 8, run on 16 and 256 cores, respectively.  The
  wall clock time includes the whole run of the program beginning to end (file
  I/O disabled).  The adaptive runs converge to a minimum level of 3.}
\label{tab:swirlwall}
\end{table}
Next, we examine how we can reduce the overall wall clock time of a simulation
by switching from uniformly refined meshes to adaptive refinement with the same maximum level,
effectively coarsening where high resolution is not beneficial.  For an example
with a sharp front as depicted in \Fig{results2d} we expect considerable
savings by AMR, which we confirm in \Tab{swirlwall}.  The wall clock times can
be reduced by up to a factor of 50 by combining AMR with subcycling.  The
adaptive runs are still  load balanced overall, which we infer from
the linear scaling of the time integration alone (not shown) and from the fact
that the weighted partition does not influence the run times.  There is the
tradeoff however, at least in the current implementation, that the adaptive
runs suffer from the loss in scalability associated with the ghost patch exchanges
described earlier. Currently, ghost patch exchanges are made across all levels, even
if only two levels are involved in the exchange.

\begin{figure}
  \begin{center}
  \begin{minipage}{.4\columnwidth}
    \begin{tabular}{cc}
      timing table for sphere & 123, \\
    \end{tabular}
  \end{minipage}
  \hspace{.05\columnwidth}
  \includegraphics[width=.4\columnwidth]{../figs_misc/sphere_partition5}
\end{center}
\caption{Partition of the spherical Example 2 between five MPI processors, indicated
  by the color scale, at simulated time $T = 1$.  The partition shows the mesh equator,
  while the refinement is invariant under the number of processors and the partition
and follows the front, which is not aligned with the equator.}
\label{fig:spherical}
\end{figure}
To demonstrate a multiblock connectivity, we include results for Example 2 in
\Fig{spherical}.

>>>>

\begin{abstract}
We present a new hybrid paradigm for parallel adaptive mesh refinement (AMR)
that combines the scalability and lightweight architecture of tree-based AMR
with the computational efficiency of patch-based solvers for hyperbolic
conservation laws.  The key idea is to interpret each leaf of the AMR hierarchy
as one uniform compute patch in $\sR^d$ with $m^d$ degrees of freedom, where $m$ is
customarily between 8 and 32.  Thus, computation on each patch can be optimized
for speed, while we inherit the flexibility of adaptive meshes.  In our work we
choose to integrate with the \pforest AMR library since it allows us to compose
the mesh from multiple mapped octrees and enables the cubed sphere and other
nontrivial multiblock geometries.  We describe aspects of the parallel
implementation and close with scalings for both MPI-only and OpenMP/MPI hybrid
runs, where the largest MPI-only run uses 16,384 CPU cores.
\end{abstract}

@@

With MPI parallelism, neighbor patches can be either local or assigned to a
different process (ghost patches).  Since this must make no difference
numerically, we need to ensure that the values of parallel neighbor patches are current
whenever they are accessed.  To this end, we allocate storage for a layer of
ghost patches in \forestclaw, which we can pass to a general-purpose \pforest
routine that communicates local leaf data to all processes that view a
particular leaf as a ghost.  If we call this routine before we go through the
local neighbor interactions, we can handle ghost values implicitly by neighbor
lookups without querying if they are local or remote.

@@

The threaded parallelism over the degrees of freedom of a patch can be handled
by \forestclaw alone without the need to involve \pforest.  For instance, a
many-core implementation, such as \manyclaw \cite{manyclaw}, can be used for
the integration of the hyperbolic system on a leaf-patch thereby allowing for
hybrid parallelism.  The design of leaf-patches then can enable efficient
management of data local to many-core architectures (for example the Xeon Phi
coprocessor) and the host.

@@

In both sets of numerical results, we solve a scalar advection
equation,
\begin{equation}
q_t + ({\bf u}\cdot \nabla) q = 0 ,
\end{equation}
where the velocity field ${\bf u} = (u(\xi,\eta,t), v(\xi,\eta,t)$ is
a prescribed function of computational coordinates.  The relevant
numerical parameters that are set in each case include the size of the
patch on each leaf ($m=8$, $16$, or $32$), and the minimum and maximum
refinement levels, which in turn fix minimum and maximum effective mesh
resolutions.

@@

\begin{description}
\item{\bf Example 1:} An initial concentration field $q$
is set to 0 in the left half of a computational square and to 1 in the
right half.  A time dependent flow field is prescribed that distorts
the interface between the 0 and 1 concentration values.  
\Fig{results2d} shows the results at an intermediate time step, where
the minimum and maximum levels of refinement are set to 3 and 6, respectively.
This results in a minimum resolution of $64 \times 64$ and a maximum
resolution of $512 \times 512$.

@@

\item{\bf Example 2:} We demonstrate the
multiblock functionality of \forestclaw by considering flow on a
sphere.  The sphere mapping we use consists of two quadtrees, each
defined in the computational space $[-1,1] \times [-1,1]$
\cite{ca-he-le:2008, be-ca-he-le:2009}.  Each quadtree is mapped to cover one
hemisphere.  The initial condition is $q = 0$ in one half of
the sphere, and $q = 1$ in the other half, where the halves are not
necessarily aligned with the equator of the mapping.  The flow field
${u}$ simulates rigid body rotation.  We show the results at an intermediate
time in \Fig{results2d}.
\end{description}

\begin{figure}
\begin{center}
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_advance_strong-crop}
  \\[1ex]
  \includegraphics[angle=-90,width=.8\columnwidth]{pdf/uni_exchange_strong-crop}
\end{center}
\caption{Strong scaling of MPI parallelism for advection on a
  uniformly refined mesh.  Top: run time of time integration.  Bottom:
  time spent in local and parallel neighbor exchanges.  The number of
  patches is $2^{2 \ell}$ at level $\ell$.  The flat lines for the
  smaller runs are caused by having fewer patches than MPI processes,
  leaving some of the processors idle.  For each data point we use the
  time required by the slowest processor to verify that the load is
  balanced equally.
%  all processes. \comment{This last
%    sentence is a bit unclear.  What do you mean exactly?}
}
\label{fig:uniscale}
\end{figure}%
We begin our analysis of Example 1 with uniformly-refined mesh experiments run on TACC's
Stampede supercomputer at different levels of resolution.
We use 1 to 16 MPI processes on one 16-core compute node, obtaining a strong
scaling efficiency for the time integration between 97\%--103\% depending on
the resolution (from levels 1 and 8).  Then, we examine processor counts from
16 to 256 using 16 cores per node.  The scaling behavior remains linear for
the time integration itself, and mostly linear for the local and parallel
neighbor exchanges which require about 10\% of the time integration run time;
see \Fig{uniscale}.

We have examined weak scalability as well by using four times as many processes
for each level increase, from level 6 at 64 cores to level 11 at 16,384.  Up to
4,096 cores, the exchange times stay consistently below 16\% of the time
integration and the time in seconds per time integration step varies between
.0578 and .0602, which yields a parallel efficiency of 96\%.  For the
data point at 16,384 cores, the efficiency is still 82\%.

\begin{table}
  \begin{center}
    \begin{tabular}{cccc|rr}
      \multicolumn{4}{c|}{strategy} & \multicolumn{2}{c}{wall clock time} \\
      mesh & remesh & partition & time step & $P = 16$ & $P = 256$ \\
      \hline
      uniform & none       & by count  & global   & 3961.  & 256.  \\
      \hline
          AMR & every step & by count  & global   &  252.  &  54.6 \\
          AMR & every 4    & by count  & global   &  178.  &  39.7 \\
      \hline
          AMR & every step & by count  & subcycle &   99.9 &  17.3 \\
          AMR & every 4    & by count  & subcycle &   87.2 &  14.0 \\
      \hline
          AMR & every step & by weight & subcycle &   95.7 &  18.2 \\
          AMR & every 4    & by weight & subcycle &   84.4 & 14.2 \\
      \hline
    \end{tabular}
  \end{center}
\caption{Comparison of different meshing strategies for the advection example
  with a fixed maximum level 8, run on 16 and 256 cores, respectively.  The
  wall clock time includes the whole run of the program beginning to end (file
  I/O disabled).  The adaptive runs converge to a minimum level of 3.}
\label{tab:swirlwall}
\end{table}
Next, we examine how we can reduce the overall wall clock time of a simulation
by switching from uniformly refined meshes to adaptive refinement with the same maximum level,
effectively coarsening where high resolution is not needed.  For an example
with a sharp front as depicted in \Fig{results2d} we expect considerable
savings by AMR, which we confirm in \Tab{swirlwall}.  The wall clock times can
be reduced by up to a factor of 50 by combining AMR with subcycling.  The
adaptive runs are still  load balanced overall, which we infer from
the linear scaling of the time integration alone (not shown) and from the fact
that the weighted partition does not influence the run times.  There is the
tradeoff however, at least in the current implementation, that the adaptive
runs exhibit reduced scalability associated with the ghost patch exchanges
described earlier.  Presently, ghost patch exchanges are made across all levels, even
if only two levels are involved in the exchange.  One enhancement of the code that is
currently being developed is to rewrite the time stepping loop with the
objective of minimizing the parallel neighbor exchanges.

\begin{figure}
  \begin{center}
  \begin{minipage}{.4\columnwidth}
    \begin{tabular}{cc}
      timing table for sphere & 123, \\
    \end{tabular}
  \end{minipage}
  \hspace{.05\columnwidth}
  \includegraphics[width=.4\columnwidth]{../figs_misc/sphere_partition5}
\end{center}
\caption{Partition of the spherical Example 2 between five MPI processors, indicated
  by the color scale, at simulated time $T = 1$.  The partition shows the mesh equator,
  while the refinement is invariant under the number of processors and the partition
and follows the front, which is not aligned with the equator.}
\label{fig:spherical}
\end{figure}
To demonstrate a multiblock connectivity, we include results for Example 2 in
\Fig{spherical}, together with a strong scaling table that shows that the spherical
example scales as well.
